Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-04-03 submission/manuscript.Rmd
Add NAL comments

The gut microbiome is thought to have a role in the development of colorectal cancer by protecting against and exacerbating inflammation. Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals that was previously subjected to 16S rRNA gene and metagenomic shotgun sequencing. We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no meaningful differences in their concentration and disease status. When we incorporated these concentrations into Random Forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated Random Forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although it is likely that bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited value.

@@

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental and behavioral factors and diet [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to create mutations, induce inflammation, and accelerate tumorigenesis. Numerous studies in murine models have supported this model [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016]. Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018]. These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988]. Based on these limited results, one would hypothesize increasing SCFA concentrations could limit tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.

@@

**Figure S1. Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the four SCFAs as additional features. Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.

>>>>

Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation. Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data. We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no meaningful association between SCFA concentration and disease status. When we incorporated these concentrations into Random Forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated Random Forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although other bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited predictive power.

@@

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental and behavioral factors and diet [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to cause mutations, induce inflammation, and accelerate tumorigenesis [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016]. Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018]. These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988]. In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.

@@

**Figure S1. Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the three SCFAs as additional features. Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.
